The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-19 manuscript/manuscript.Rnw
use PSG and PAG acronyms; update introduction

% We identified significant shifts in gene expression along specific
% evolutionary lineages using a model of character evolution. Genes that
% showed significant elevation in expression along the human lineage
% ($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
% angiogenesis, tube morphogenesis, cell migration, and regulation of
% inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
% showed exceptionally high expression in humans; these genes are
% implicated in preeclampsia and may indicate that the risk of
% preeclampsia recently emerged on the human lineage.

@@

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}.

@@

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., humans, mice, cattle) have
revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in mouse and
cow, pregnancy-specific glycoproteins (PSGs) in humans and mouse,
pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
the growth hormone and chorionic gonadotropin beta genes in hominoid
primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between humans and mice
\cite{Su.ea2002:Large-scaleanalysisofhuman}, and the recruitment of
\textit{SIGLEC6} to placental expression in humans differs from both
chimpanzees and mice
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

One explanation for the divergent paths of the placental interface is
the distinct evolutionary interplay between maternal and fetal
strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
The maternal strategy is to requisition an adequate degree of
nutrients to the fetus without risking the health of the mother, while
the fetal strategy via paternal genes is to extract as many resources
from the mother as possible. Therefore, an interplay between these
extremes can set up an intricate “arms race”, involving the imprinting
of parental alleles, production of new genes, and enlistment of
existing genes to expression at the maternal-fetal interface
\cite{Haig1993:Geneticconflictsinhuman}.

One symptom of this maternal-fetal conflict is its potential affect on
human health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, humans possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of preeclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in chimpanzees
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
gorillas \cite{Thornton.ea1992:Convulsionsinpregnancyin}, preeclampsia
has a far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Chimpanzees
and gorillas also have deeply invasive trophoblasts, although the
progression of invasion in great apes may be slower than humans
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \bonobo (the sister
species to the chimpanzee), \smonkey, \mouse, \galili, \carmeli, \dog,
\cow, \elephant, \armadillo, and \opossum, representing marsupials as
well as three major branches of eutherian mammals (Euarchontoglires,
Laurasiatheria, Afrotheria). These placentas also represent a wide
array of placental morphologies as well as traits such as number and
size of offspring and length of gestation
\cref{tab:gestational_periods}. We use this data, along with existing
RNA-Seq data from \mouse \cite{Wang.ea2011:surveyfornovelimprinted},
\human \cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse \cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify human specific differences in gene
expression that may be related to obstetrical disorders such as
preeclampsia.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven pregnancy-specific beta-1-glycoprotein genes are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
preeclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Preeclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
preeclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
preeclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. Although there are speculative cases of
eclampsia in great apes
\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is possible
that a comparative evolutionary approach to gene expression can
identify genes that have divergent expression patterns from other
mammals as well as association with preeclampsia.

@@

% Many genes that both show expression shifts on the human lineage and
% have very high expression (FPKM>200) are directly related to the
% highly invasive nature of the human placenta. For example, GPC3,
% glypican 3, is an imprinted gene and cell growth regulator, associated
% with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
% intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
% 1996). GPC3 has the highest expression in humans (Table 2), but is
% also expressed in other mammals to a lesser degree, especially those
% with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
% spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
% (Supplemental Tables X-XX). ADAM12 increases the availability of
% insulin-like growth factors that lead to the promotion of cell growth
% and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
% found evidence of adaptive evolution in ADAM12 on the primate lineage,
% and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
% monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
% lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
% associated with trophoblast invasion as well (Bilban et al., 2010;
% Yuen et al., 2011). In addition, one other gene, KISS1, shows
% exceptional abundance in the human placenta (Table 2); however due to
% its absence in elephant Ensmebl gene annotations, it was not part of
% the 1:1 orthologues examined in the comparative study. Here, it is
% only expressed in the term placentas of humans (Table 2) and bonobos
% (\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
% in the regulation of trophoblast invasion, and shows highest
% expression in the first trimester of pregnancy (Bilban et al., 2004;
% Reynolds et al., 2009).
% 
% In addition, several other genes that show expression shifts on the
% human lineage also have major effects on placental function. Genes
% such as TXK and EBI3 are intimately involved in the immune system in
% the placenta. TXK is involved in natural killer cell differentiation
% (Bezman et al., 2012). EBI3 is implicated in natural killer cell
% function an maternal-fetal immunomodulation in the syncytiotrophoblast
% (Devergne et al., 2001). KYNU and KMO are interacting genes that are
% involved in tryptophan metabolism, which may be involved in
% immunosupression and the prevention of fetal rejection (Takikawa,
% 2005). THSD7A is a thrombospondin, which inhibits endothelial cell
% migration and tube morphogenesis (Wang et al., 2010).
% 
% Many genes that show evidence of a shift on the human lineage show
% evidence of implication in preeclampsia, including ADAM12 (Gack et
% al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
% Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
% al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
% MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
% In addition KISS1 is also involved in preeclampsia as well (Sado et
% al., 2011; Qiao et al, 2012). Due to the association of many genes
% that show human-specific shifts as well as high expression, we
% conclude that preeclampsia is possibly a symptom of recent evolution
% of gene expression on the human lineage. Although many genes such as
% GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
% this expression has intensified on the human lineage. Sequencing of
% more placentas of nonhuman primates, including other individuals of
% the genus Pan will help to further characterize this transition.

>>>>

% We identified significant shifts in gene expression along specific
% evolutionary lineages using a model of character evolution. Genes that
% showed significant elevation in expression along the human lineage
% ($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
% angiogenesis, tube morphogenesis, cell migration, and regulation of
% inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
% showed exceptionally high expression in humans; these genes are
% implicated in pre-eclampsia and may indicate that the risk of
% pre-eclampsia recently emerged on the human lineage.

@@

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

@@

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \human, \mouse, \cow) have
revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\carmeli, \galili, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \elephant, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{P. troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

One explanation for the divergent paths of the placental interface is
the distinct evolutionary interplay between maternal and fetal
strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
The maternal strategy is to requisition an adequate degree of
nutrients to the fetus without risking the health of the mother, while
the fetal strategy via paternal genes is to extract as many resources
from the mother as possible. Therefore, an interplay between these
extremes can set up an intricate “arms race”, involving the imprinting
of parental alleles, expansion of gene families, and enlistment of
existing genes to expression at the maternal-fetal interface
\cite{Haig1993:Geneticconflictsinhuman}.

One symptom of this maternal-fetal conflict is its potential affect on
\human health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, \human possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of pre-eclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in \textit{P. troglodytes}
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
\textit{Gorilla gorilla}
\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
far higher incidence in \human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
  troglodytes} and \textit{G. gorilla} also have deeply invasive
trophoblasts, although the progression of invasion in great apes may
be slower than \human
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \bonobo (the sister
species to \textit{P. troglodytes}), \smonkey, \mouse, \galili,
\carmeli, \dog, \cow, \elephant, \armadillo, and \opossum. These
species represent three major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria) as well as marsupials.
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation (\cref{tab:gestational_periods}). We use this data, along with
existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify \human specific differences in gene
expression that may be related to obstetrical disorders such as
pre-eclampsia.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Preeclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. Although there are speculative cases of
eclampsia in great apes
\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is possible
that a comparative evolutionary approach to gene expression can
identify genes that have divergent expression patterns from other
mammals as well as association with pre-eclampsia.

@@

% Many genes that both show expression shifts on the human lineage and
% have very high expression (FPKM>200) are directly related to the
% highly invasive nature of the human placenta. For example, GPC3,
% glypican 3, is an imprinted gene and cell growth regulator, associated
% with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
% intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
% 1996). GPC3 has the highest expression in humans (Table 2), but is
% also expressed in other mammals to a lesser degree, especially those
% with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
% spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
% (Supplemental Tables X-XX). ADAM12 increases the availability of
% insulin-like growth factors that lead to the promotion of cell growth
% and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
% found evidence of adaptive evolution in ADAM12 on the primate lineage,
% and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
% monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
% lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
% associated with trophoblast invasion as well (Bilban et al., 2010;
% Yuen et al., 2011). In addition, one other gene, KISS1, shows
% exceptional abundance in the human placenta (Table 2); however due to
% its absence in elephant Ensmebl gene annotations, it was not part of
% the 1:1 orthologues examined in the comparative study. Here, it is
% only expressed in the term placentas of humans (Table 2) and bonobos
% (\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
% in the regulation of trophoblast invasion, and shows highest
% expression in the first trimester of pregnancy (Bilban et al., 2004;
% Reynolds et al., 2009).
% 
% In addition, several other genes that show expression shifts on the
% human lineage also have major effects on placental function. Genes
% such as TXK and EBI3 are intimately involved in the immune system in
% the placenta. TXK is involved in natural killer cell differentiation
% (Bezman et al., 2012). EBI3 is implicated in natural killer cell
% function an maternal-fetal immunomodulation in the syncytiotrophoblast
% (Devergne et al., 2001). KYNU and KMO are interacting genes that are
% involved in tryptophan metabolism, which may be involved in
% immunosupression and the prevention of fetal rejection (Takikawa,
% 2005). THSD7A is a thrombospondin, which inhibits endothelial cell
% migration and tube morphogenesis (Wang et al., 2010).
% 
% Many genes that show evidence of a shift on the human lineage show
% evidence of implication in pre-eclampsia, including ADAM12 (Gack et
% al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
% Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
% al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
% MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
% In addition KISS1 is also involved in pre-eclampsia as well (Sado et
% al., 2011; Qiao et al, 2012). Due to the association of many genes
% that show human-specific shifts as well as high expression, we
% conclude that pre-eclampsia is possibly a symptom of recent evolution
% of gene expression on the human lineage. Although many genes such as
% GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
% this expression has intensified on the human lineage. Sequencing of
% more placentas of nonhuman primates, including other individuals of
% the genus Pan will help to further characterize this transition.
